Estimating Fundamental Stellar Properties using PHOENIX Grid Interpolators

### 2023-11-28 paper/paper2/ms.tex
Big paper work

\usepackage{amsmath}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{enumitem}
% Temporary packages
\usepackage{mdframed}
\usepackage{verbatim}

\def\Teff{T_{\rm eff}}
\def\vsini{v\sin{i}}
\def\kmps{\mathrm{km}\;\mathrm{s}^{-1}}

\begin{document}
\title{Blas\'e3D}
\shorttitle{\emph{blas\'e3D}}

@@

@@

\begin{mdframed}
    \textbf{The PHOENIX grid subset}

\begin{mdframed}
    \textbf{Cloning with blas\'e}

    - Per-grid point optimization procedure
    - Pretrained model caching
    - Output storage
    \\---------------\\
    Blas\'e's algorithm is designed to take a spectrum as input, and outputs two products:
    1. A list of detected spectral line properties, all assumed to be Voigt profiles, and 2.
    A cloned spectrum, which is the aggregation of all of those detected lines into a spectrum. We
    don't use the cloned spectrum from this process, and only focus on the line properties that are output.
    These two products are stored as a .pt file, which can then be loaded in in order to 
    analyze all the lines in bulk. Each grid point is processed in this way, and the resulting .pt files'
    line properties are aggregated into a single DataFrame, which can then be queried by $T_eff$, $\log(g)$, $Z$,
    in addition to the actual line in question, identified by its center $\mu_j$.
\end{mdframed}

\begin{mdframed}
    \textbf{Open Source Availability}
    \\--------------\\
    The PHOENIX model grid is available open-source at $xx$, and can be interfaced with gollum,
    also available open-source at $xx$. Blas\'e is also available open-source at $xx$, with this work
    existing as a fork named blas\'e3D, available at $xx$.
\end{mdframed}

@@

\begin{mdframed}
    \textbf{Line Recognition}
    - Need: Identifying unique lines
    - Anticipated friction points: line centroids drift
    - Strategy: (Pre-shift centers)
    \textcolor{lightgray}{\blindtext}
\end{mdframed}

\begin{mdframed}
    \textbf{Line property bulk trends across the grid}
    \textcolor{lightgray}{\blindtext}
\end{mdframed}

\begin{mdframed}
    \textbf{Figure: Heatmap $T_\mathrm{eff}$ vs $\log{g}$ for a single line $2\times2$ panels for $\sigma$, $\gamma$, $A$, $\lambda$ }
\end{mdframed}

\begin{mdframed}
    \textbf{Anomalies in Heatmaps}
    \textcolor{lightgray}{\blindtext}
\end{mdframed}

@@

\begin{mdframed}
    \textbf{Conceivable Extensions}
    \\-------------\\
    Future studies can extend this work in several ways:
    \begin{enumerate}[label=-]
        \item Extending the data from just the PHOENIX Grid to additional synthethic spectral
              model grids, such as the Sonora substellar model grids or the COOLTLUSTY substellar model grid.
    \end{enumerate}
\end{mdframed}

>>>>

\usepackage{amsmath}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{hyperref}
% Temporary packages
\usepackage{mdframed}
\usepackage{verbatim}

\def\Teff{T_{\rm eff}}
\def\logg{\log(g)}
\def\Z{\mathrm{[Fe/H]}}
\def\vsini{v\sin(i)}
\def\kmps{\mathrm{km}\;\mathrm{s}^{-1}}

\begin{document}
\title{Estimating Fundamental Stellar Properties using PHOENIX Grid Interpolators}

@@

Precomputed synthetic stellar spectral models represent a distillation of all available physics
and chemistry theory into effortlessly-reproducible comparisons for spectral observations.  
Practitioners may simply download a file and overplot it against an observed spectrum, relying on 
human pattern recognition to obtain insights on fundamental stellar properties. But such exercises 
typically fall short. Inevitably, individual line depth and shape predictions do not match the 
observed spectrum. \citet{2022ApJ...941..200G} introduced a two-stage interpretable machine learning (ML) 
technique to alleviate some of these challenges. First, cloning the lines in a precomputed spectrum 
with fine-tuned Voigt profiles, and then warping their line properties in response to data using 
the power of automatic differentiation. This design satisfied a common use case, the desire to 
obtain a semi-empirical super-resolution stellar template for the observed star: the pristine 
$R=\frac{\lambda}{\delta \lambda}\sim1,000,000$ emergent spectrum coming from a patch of the star, 
having accounted for rotational broadening, instrumental resolution, and telluric artifacts. 
We called this algorithm \emph{blas\'e}, available with the Python package \texttt{blase}.

This capability alone was useful for Extreme Precision Radial Velocity (EPRV), but a key limitation 
of \emph{blas\'e} was its requirement to pre-select a grid point as the initial starting point. The code
introduced in this study bypasses this requirement by comparing spectral reconstructions to an input
spectrum in order to automatically ascertain the fundamental stellar parameters.

@@

For the purposes of this study, we use a slice of the PHOENIX grid where the alpha abundance
is set to zero. Within this slice, we allow the full range of $\Teff, \logg, \Z$. The specific
sampling of the grid points is shown in Figure $XX$. Our approach could feasibly be extended to
include alpha abundances, but given that the \texttt{gollum} interface does not yet support the access
of spectra with nonzero alpha abundances, as well as bearing a significant increase in computational 
impact, we leave this for future work.

The \texttt{blase} code takes in any spectrum as input, and outputs a cloned
version of that spectrum. However, there is a hidden product to this
procedure, the state dictionary of the model itself, stored as a \texttt{.pt} file. In this lies metadata about
where spectral lines were identified by \texttt{blase}, as well as those lines'
specific properties, namely amplitude, Gaussian shape, and Lorentzian shape.
In addition, \texttt{blase} also allows the line centers to shift slightly during
cloning, meaning that in addition to the initial line centers, we 
also have access to the shifted line centers. This hidden metadata is 
what formed the backbone for this study and allowed us to retrieve line-by-line data
that could then be interpolated. We input the PHOENIX grid spectra into \texttt{blase}
and accessed the state dictionaries of each clone. Once this data was retrieved from the state dictionaries
across the grid, it was then aggregated into a single \texttt{pandas} DataFrame, enabling
querying of the data by any property, be it a fundamental stellar parameter or the output
line properties.

All data and code used in this study is available online and is open source. The PHOENIX grid itself takes quite a while to download
due to server bandwidth limitations; this is something that \texttt{gollum} aims to
improve upon by implementing both caching behavior and underlying data structure compression.
\texttt{gollum} itself is available on GitHub and is pip installable, although this
method lags in features significantly compared to the bleeding edge code from GitHub.
\texttt{blase} is the main library of note here, also on GitHub, while the
materials related to this study are currently residing as a fork named \texttt{blase3D}. When polished and complete,
the fork is set to merge into \texttt{blase} itself. Links to all of these are provided in the appendix.

@@

We have established that the \texttt{blase} code is capable of cloning a spectrum and
returning a set of line properties for each line. In order to visualize line-by-line
properties across the grid, we need to be able to identify unique lines. This begs the question:
what property lets us uniquely identify a spectral line? We have two possible solutions.
First, we consider the shifted line centers. However, this shifting comes from \texttt{blase}'s
emulating physics from its training dataset, meaning that the same line at two grid points could
be shifted differently. Could we then round or truncate the wavelength coordinate of the line centers?
The answer is no, because these shifts vary in precision, and some lines are close enough to begin with
that what should have been two different lines get identified as the same line. This, obviously is incorrect;
one spectral line is just that, one spectral line. So we now consider the original unshifted line centers.
These are the same across the grid, and there is no issue with unintentionally binning lines together
thanks to a uniform precision. This means that we now use the unshifted line centers as the unique
identifier for lines, which vary across the grid in their properties, namely amplitude, Gaussian shape,
and Lorentzian shape.

The next thing that we want to do is to visualize the trends of line properties over
the PHOENIX grid. The best way to visualize this is with heatmaps. This means that for
each line, for each line property, we have one heatmap, where the axes are the fundamental
parameters, and the `temperature' is the value of the line property. As you can imagine, this
is a lot of heatmaps, each of which is incredibly hard to visualize due to the dimensionality.
So, we slice the heatmaps at solar metallicity to reduce them to 2D, and then we can easily 
visualize the trend of any one line property of any one line across the grid on demand.

\begin{mdframed}
    \textbf{Figure: Heatmap $\Teff$ vs $\logg$ for a single line $2\times2$ panels for $\sigma$, $\gamma$, $A$, $\lambda$ }
\end{mdframed}

While we see heatmaps for lines that have values at all grid points, this is only true for
a select few spectral lines. The vast majority of lines do not seem to show up at certain
grid points, meaning that our heatmaps are sparse. Initially this seems like it would not be an 
issue; when reconstructing, simply retain knowledge of which points were missing and override
the interpolated value with a missing value. Now, in addition to an
interpolator, a sequence of grid points with missing values must be stored for each line just to
override the interpolator. This increases the storage cost of the model. In addition to this, we would
not be able to deal with points lying on the boundary between missing and visible lines. So, we need
to process these missing lines. Since the line properties are on a logarithmic scale,
a line with an amplitude of 0 would have a log amplitude of $-\infty$. This means that we should just
set all missing lines to have log amplitudes of $-\infty$, right? Close, however interpolation
algorithms absolutely despise infinities. So, we set the log amplitude of missing lines to be
$-1000$, which is a large enough negative number that it will effectively evaluate to zero without
causing problems with the interpolation process. We do not really have to worry about the Gaussian
and Lorentzian shape parameters; no matter how `wrong' their interpolated values are, an amplitude of
zero will override that and make the line disappear. This way, we have effectively dealt with
pathological behavior in such a way that interpolation returns a sensible value without sending
the storage impact through the roof.

@@

Future studies can expand upon this work in several ways:
\begin{enumerate}[label=-]
    \item Extending the data from just the PHOENIX grid to additional synthetic spectral
          model grids, such as the Sonora substellar model grids or the COOLTLUSTY substellar model grid.
    \item Expanding the section of the PHOENIX grid on which \texttt{blase} was run to include all available alpha abundances, encompassing all of PHOENIX.
    \item Using more advanced regressors with ML techniques to predict line properties instead of using interpolation to brute-force the model predictions.
    \item Extending support to accommodate large-bandwidth spectra across the full electromagnetic spectrum.
    \item Identifying spectral lines using species-based line lists rather than fitting Voigt profiles whenever possible.
\end{enumerate}
